---
title: "Activity Monitoring Analysis"
---

##Loading and preprocessing the data

The first step is to find the file online and load it to further analyze.

```{r, echo=TRUE}
# check if a data folder exists
if (!file.exists("data")) {dir.create("data")}

# file URL and destination file
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile <- "./data/activity.zip"

# download the file and note the time
download.file(fileUrl, destfile = destfile)
dateDownloaded <- date()

# Read out the containing csv file
data <- read.csv(unz("./data/activity.zip", "activity.csv"))
```

##What is mean total number of steps taken per day?

Next, the data is divided up into date, steps, and interval.

```{r,echo=TRUE}
# assign variables to the columns
steps <- data$steps
date <- data$date
interval <- data$interval

# filter to isolate non-NA values
filter <- !is.na(steps)

# apply the steps filter to the date vector
filter_steps <- steps[filter]
filter_date <- date[filter]
```

Another vector was established to determine the total steps each day.

```{r,echo=TRUE}
# create a factor vector for the non-NA days
days_factor <- factor(filter_date)

# get the total number of steps for each day
total_steps <- tapply(filter_steps, days_factor, FUN = sum)

# plot a histogram of the total number of steps taken each day
hist(total_steps, breaks = 10, 
          xlab = "Total number of steps per day", 
          main = "Distribution of total steps per day", 
          col = "red", 
          type = "count")
```

The mean and median were also calculated.

```{r,echo=TRUE}
original_mean <- mean(total_steps)
```

##[1] 10766

```{r,echo=TRUE}
original_median <- median(total_steps)
```

##[1] 10765

## What is the average daily activity pattern?

To determine the average daily acvitivy pattern, a vector was established to calculate the average steps.

```{r,echo=TRUE}
# create a factor vector for the time intervals
interval_factor <- factor(interval)
levels <- nlevels(interval_factor)
interval_factor <- factor(interval)[1:levels]

# calculate the average number of steps for each 5 minute period
average_steps <- tapply(steps, factor(interval), FUN = mean, na.rm = TRUE)
average_steps <- sapply(average_steps, simplify = array, round, 2)

scales=list( x=list(at = seq(0, 2400, 200)))     
   
# plot the time series
plot(as.numeric(average_steps) ~ interval[1:288], 
       type = "l", 
       xlab = "Time interval",
       ylab = "Average steps", 
       main = "Time series - average steps vs time interval", 
       scales = scales)
```

Using average steps and time, a data fram was created to sort steps and time.

```{r,echo=TRUE}
# create a data frame of average steps and time interval
df_steps_interval <- data.frame(interval_factor, average_steps)

# sort df to get the row with the maximum amount of average steps
df_steps_interval <- df_steps_interval[order(df_steps_interval$average_steps, 
                                             decreasing = TRUE),]

# the first row contains the relevant time interval
time_interval_max <- df_steps_interval$interval_factor[1]

# convert the factor to a character and then to numeric
time_interval_max <- as.numeric(as.character(time_interval_max))
```

##[1] 835

## Imputing missing values

The number of missing values was determine by the following code:

```{r,echo=TRUE}
# number of NA values in original dataset
length(steps[is.na(steps)])
```

## [1] 2304

Then those values were assumed new ones built on the analogous 5 minute interval for the average steps.

```{r,echo=TRUE}
# take a copy of the original steps vector
new_steps <- steps

# fill in each NA value by taking the average for that time interval
for (i in which(sapply(new_steps, is.na))) {
  
  # set the value to the equivalent value in the average vector
  if (i <= 288){
    new_steps[i] <- average_steps[i]
  } 
  
  # wrap around 288 (avg time only has 24 hours of data) and add one because 
  # R is non-zero index
  else{
    j <- i%%288 + 1
    new_steps[i] <- average_steps[j]
  }
}
```

The sum was then determined and a histogram was created to display the results.

```{r,echo=TRUE}
# create a factor vector for all of the days
new_days_factor <- factor(new_steps)

# get the total number of steps for each day
new_total_steps <- tapply(new_steps, new_days_factor, FUN = sum)

# plot a histogram of the total number of steps taken each day
hist(new_total_steps, breaks = 10, 
          xlab = "Total number of steps per day", 
          main = "Total steps per day after imputted values", 
          col = "red",
          type = "count")
```

The mean and median were then calculated

```{r,echo=TRUE}
new_mean <- mean(new_total_steps)
```

## [1] 755.7

```{r,echo=TRUE}
new_median <- median(new_total_steps)
```

## [1] 612

## Are there differences in activity patterns between weekdays and weekends?

Build a function to establish whether the date is, in fact, a weekday.

```{r,echo=TRUE}
week_day <- function(date_val) {
    wd <- weekdays(as.Date(date_val, '%Y-%m-%d'))
    if  (!(wd == 'Saturday' || wd == 'Sunday')) {
        x <- 'Weekday'
    } else {
        x <- 'Weekend'
    }
    x
}
```

Then, the function is used to determine a new variable.

```{r, echo=TRUE}
# Apply the week_day function and add a new column to activity dataset
data$day_type <- as.factor(sapply(data$date, week_day))

#load the ggplot library
library(ggplot2)

# Create the aggregated data frame by intervals and day_type
steps_per_day_impute <- aggregate(steps ~ interval+day_type, data, mean)

# Create the plot
plt <- ggplot(steps_per_day_impute, aes(interval, steps)) +
    geom_line(stat = "identity", aes(colour = day_type)) +
    theme_gray() +
    facet_grid(day_type ~ ., scales="fixed", space="fixed") +
    labs(x="Time", y=expression("Steps")) +
    ggtitle("Number of steps vs. Time interval") + theme(legend.position="none")
print(plt)
```